Using \LaTeX\ to submit manuscripts to peer reviewed journals.

### 2014-03-13 docs/thesis/chap-cern_math_explorer.tex
proofreading fixes

\subsubsection{Unicode Normalizing} 
The first issue we discovered is that because of the wide range of different characters to represent the possible mathematical symbols, there is need for an automatic way to match different possible variations of a single character. An example of this situation is the specification of an natural number, sometimes denoted by the letter N. In this case, at least three different common representations were identified: $ N = 1 $ (Using character with unicode code point 0x004e), $\mathcal{N} = 1$ (0xd835 0xdc41 or in latex expressed as \lstinline|\mathcal{N}|) and $\mathbb{N} = 1$ (0x2115 or in latex expressed as \lstinline|\mathbb{N}|). \\
  To address this situation, we rely on the Unicode equivalence framework. In the unicode specification, sometimes the same character has two or more different code-points because of backwards compatibility with other encoding standards or because the same character has different essential meanings. Another group of equal characters with different code-points are pre-combined characters such as the Latin letter \~{N} which has Unicode point 0x00D1 and also is the sequence of code-points 0x004E (Latin letter N) and 0x0303 (the combining character tilde). Taking this sample into account, Unicode defines the character composition and decomposition operations as replacing the decomposed representation with the pre-composed one and vice-versa. Finally, Unicode defines two types of equivalences between characters: Canonical and compatible. Canonical equivalent characters are assumed to have the same appearance and meaning. Compatible ones are allowed to have slightly different graphical representations but the same meaning in some contexts. One example of compatible equivalent characters is the Roman number 12 {\unicodefontⅫ} with code-point 0x216b and the sequence of characters XII with codepoints 0x0058 0x0049 0x0049. Taking into account the composition or decomposition of characters and the canonical or compatible equivalences, Unicode specifies 4 different forms of a given character:
  \begin{itemize}
  \item NFD: Normalization Form Canonical Decomposition
  \item NFC: Normalization Form Canonical Composition
  \item NFKD: Normalization Form Compatibility Decomposition
  \item NFKC: Normalization Form Compatibility Composition
  \end{itemize}

@@

\subsubsection{Operator grouping}
Even though the Unicode normalization step works for a big part of the cases presented in table \ref{unicode_characters}, there are still some groups of characters which would not be matched (even partially), such as some types of integrals. To address these groups, a precomputed table was created for some general groups of operators which have some similar meaning. For each token that belongs to some of this predefined groups, a second token is created identifying the group.
For example the token <mo>{\unicodefont ∮}</mo> is expanded into <mo>{\unicodefont ∮}</mo><mo>INTEGRALS</mo>, similarly the token <mo>{\unicodefont ∫}</mo> is expanded into <mo>{\unicodefont ∫}</mo><mo>INTEGRALS</mo> and then both will have a partial match in the INTEGRALS token. 
The Unicode specification does not define a strict rule to group similar characters based on their semantics. For general operators, we used the grouping provided by Xah Lee in his website\cite{math_groups}.

@@

\end{longtable}
\subsection{Recursive interpretation}
Unfortunately, Mathematica cannot always interpret correctly an input expression. In this case, it can, however interpret some sub-expressions from the original one. Our system takes advantage of this, and the first attempt is to try to interpret the original string. If this step is not successful (the result variable equals the string "{}"), we identify the occurrences of {\codefont RowBox} which identify all the subexpressions in the tree. For each of this elements we call the {\codefont MakeExpression} and if it was successful we search for structural pattern in it. A first thought would tell us that if sub-tree was interpreted successfully, there is no need to try to go further down in that sub-tree, however, even if a sub-expression is interpreted correctly, the result sometimes is not complete and processing further down, gives better results. For this reason, we do not discard any of the subexpressions and try to interpret all of them. For this same reason, we do not store the number of occurrences of a given pattern, but only whether it contains is or not.

@@

\section{Integration with CDS}
Currently the Invenio code-base is being under a deep refactoring and rewriting. CDS currently runs in the stable branch {\codefont master} while the new developments are preferred to be implemented in the new branch {\codefont pu}. 
Our system is deployed as a standalone {\codefont jar} which is added to a Solr instance. Also, some extra configurations should be added. Details on this can be found in appendix \ref{appendix-conf_files}
The communication between Solr and Invenio was through the library {\codefont solrpy}. 
Figure 5.3 presents the main components of our deployment.

>>>>

\subsubsection{Unicode Normalizing} 
The first issue we discovered is that because of the wide range of different characters to represent the possible mathematical symbols, there is need for an automatic way to match different possible variations of a single character. An example of this situation is the specification of an natural number, sometimes denoted by the letter N. In this case, at least three different common representations were identified: $ N = 1 $ (Using character with unicode code point 0x004e), $\mathcal{N} = 1$ (0xd835 0xdc41 or in latex expressed as \lstinline|\mathcal{N}|) and $\mathbb{N} = 1$ (0x2115 or in latex expressed as \lstinline|\mathbb{N}|). \\
  To address this situation, we rely on the Unicode equivalence framework. In the unicode specification, sometimes the same character has two or more different code-points because of backwards compatibility with other encoding standards or because the same character has different essential meanings. Another group of equal characters with different code-points are pre-combined characters such as the Latin letter \~{N} which has Unicode point 0x00D1 and also is the sequence of code-points 0x004E (Latin letter N) and 0x0303 (the combining character tilde). Taking this sample into account, Unicode defines the character composition and decomposition operations as replacing the decomposed representation with the pre-composed one and vice-versa. Finally, Unicode defines two types of equivalences between characters: Canonical and compatible. Canonical equivalent characters are assumed to have the same appearance and meaning. Compatible ones are allowed to have slightly different graphical representations but the same meaning in some contexts. One example of compatible equivalent characters is the Roman number 12 {\unicodefontⅫ} with code-point 0x216b and the sequence of characters XII with codepoints 0x0058 0x0049 0x0049. Taking into account the composition or decomposition of characters and the canonical or compatible equivalences, Unicode specifies 4 different forms of a given character:
  \begin{itemize}
  \itemsep0em
  \item NFD: Normalization Form Canonical Decomposition
  \item NFC: Normalization Form Canonical Composition
  \item NFKD: Normalization Form Compatibility Decomposition
  \item NFKC: Normalization Form Compatibility Composition
  \end{itemize}

@@

\subsubsection{Operator grouping}
Even though the Unicode normalization step works for a big part of the cases presented in table \ref{unicode_characters}, there are still some groups of characters which would not be matched (even partially), such as some types of integrals. To address these groups, a precomputed table was created for some general groups of operators which have some similar meaning. For each token that belongs to some of this predefined groups, a second token is created identifying the group.
For example the token <mo>{\unicodefont ∮}</mo> is expanded into <mo>{\unicodefont ∮}</mo> <mo>INTEGRALS</mo>, similarly the token <mo>{\unicodefont ∫}</mo> is expanded into <mo>{\unicodefont ∫}</mo> <mo>INTEGRALS</mo> and then both will have a partial match in the INTEGRALS token. 
The Unicode specification does not define a strict rule to group similar characters based on their semantics. For general operators, we used the grouping provided by Xah Lee in his website\cite{math_groups}.

@@

\end{longtable}
\subsection{Recursive interpretation}
Unfortunately, Mathematica cannot always interpret correctly an input expression. In this case, it can, however interpret some sub-expressions from the original one. Our system takes advantage of this, and the first attempt is to try to interpret the original string. If this step is not successful (the result variable equals the string "\{\}"), we identify the occurrences of {\codefont RowBox} which identify all the subexpressions in the tree. For each of this elements we call the {\codefont MakeExpression} and if it was successful we search for structural patterns in it. A first thought would tell us that if sub-tree was interpreted successfully, there is no need to try to go further down in that sub-tree, however, even if a sub-expression is interpreted correctly, the result sometimes is not complete and processing further down gives better results. For this reason, we do not discard any of the subexpressions and try to interpret all of them. For this same reason, we do not store the number of occurrences of a given pattern, but only whether it contains is or not.

@@

\section{Integration with CDS}
Currently the Invenio code-base is being under a deep refactoring and rewriting. CDS currently runs in the stable branch {\codefont master} while the new developments are preferred to be implemented in the new branch {\codefont pu}. 
Our system is deployed as a standalone {\codefont jar} which is added to a Solr instance. Also, some extra configurations should be added. Details on this can be found in appendix \ref{appendix-conf_files}.
The communication between Solr and Invenio was through the library {\codefont solrpy}. 
Figure 5.3 presents the main components of our deployment.

### 2014-03-13 docs/thesis/chap-related_work.tex
work on title page
proofreading fixes

\subsection{LaTEXSearch}
LaTEXSearch \url{http://latexsearch.com/} is a system developed by the scientific publisher Springer that allows to search over a database of around eight million latex snippets extracted from their publications. The system, unfortunately, is proprietary and no details are provided about the internals of the system.

\subsection{WikiMirs}
WikiMirs\cite{wikimirs} is a system that, as EgoMath, allows to find relevant Wikipedia entries by providing mathematical expression in \LaTeX. It combines the information about the semantic tree with the layout tree of a given expression and uses a similar approach to index different generalization levels.

Other systems that provide similar search functionalities are \url{Uniquation} and \url{http://www.symbolab.com/}, however detailed information about the systems are not provided. Finally Wolfram Alpha\cite{wolframalpha} is a very powerful tool that allows to find information about lots of mathematical concepts and a wide variety of other objects (Even Pokemon).

>>>>

\subsection{LaTeXSearch}
LaTeXSearch \url{http://latexsearch.com/} is a system developed by the scientific publisher Springer that allows to search over a database of around eight million latex snippets extracted from their publications. The system, unfortunately, is proprietary and no further details are provided.

\subsection{WikiMirs}
WikiMirs\cite{wikimirs} is a system that, as EgoMath, allows to find relevant Wikipedia entries by providing mathematical expression in \LaTeX. It combines the information about the semantic tree with the layout tree of a given expression and uses a similar approach to index different generalization levels. The system is available at \url{http://www.icst.pku.edu.cn/cpdp/wikimirs/} \\

Other systems that provide similar search functionalities are Uniquation(\url{http://uniquation.com}) and Symbolab(\url{http://www.symbolab.com/}), however detailed information about the systems are not provided. Finally Wolfram Alpha\cite{wolframalpha} is a very powerful tool that allows to find information about lots of mathematical concepts and a wide variety of other objects (Even Pokemon).

### 2014-03-13 docs/thesis/chap-technical_considerations.tex
proofreading fixes

OpenMath is yet another markup language to represent mathematics and can be used to compliment presentation MathML to add more semantic information. It is mainly used now for defining mathematical concepts into standard content dictionaries which can be imported. MathML is compatible with this dictionaries.

@@

\LaTeX allows the most compact representations of the equations and it is the closest language to the community. However, its processing is not straightforward because a given equation can include multiple commands from different packages. Also there are several ways to represent the same set of symbols, either using its Unicode code-point or through embedded commands. Another difficulty are custom commands that users can define in the preamble of the document. In overall, the interpretation and processing of a mathematical expression in this language would require writing a complete compiler which is not the purpose of this work. MathML offers a better compromise between the length of the representation and the easiness to be processed.

\section{Math Extraction Tools}
One of the main steps in any information retrieval task is the extraction of the searchable content from the collection of documents. At CDS the most common storage format for publications is PDF. A big part of the harvested documents in CDS comes from the ArXiv pre-prints service and from here, the source code (\LaTeX) of the submissions is available for download. Smaller portions of other documents are stored in other formats like Microsoft Word, OpenOffice Documents.
In practice, the main options from where to do the extraction from, are PDFs and \LaTeX  files. Also taking into account the previous discussion on the available format for representing mathematics, we focused our attention in tools that output MathML.  \\
A conscious exploration of available software for extracting mathematical content from this two types of documents allowed us to identify different tools with different approaches. 
\LaTeX  files are easier to process and all of the tools implement their own compiler that XML + MathML outputs instead of the typical. Since the original math equation is provided in the document, there is relatively little ambiguity in the equation, and the major issues that the systems struggle with is in the coverage of all the possible commands and packages that are available. \\
For extracting equations from PDF files, the process turns into an Optical Character Recognition (OCR). The reconstruction of the equation has to infer the mathematical structure based on physical positioning and size of the symbols, the length of the lines and so on.

>>>>

OpenMath is yet another markup language to represent mathematics and can be used to compliment presentation MathML to add more semantic information. It is mainly used now for defining mathematical concepts into standard content dictionaries which can be imported. MathML is compatible with these dictionaries.

@@

\LaTeX \ allows the most compact representations of the equations and it is the closest language to the community. However, its processing is not straightforward because a given equation can include multiple commands from different packages. Also there are several ways to represent the same set of symbols, either using its Unicode code-point or through embedded commands. Another difficulty are custom commands that users can define in the preamble of the document. In overall, the interpretation and processing of a mathematical expression in this language would require writing a complete compiler which is not the purpose of this work. MathML offers a better compromise between the length of the representation and the easiness to be processed.

\section{Math Extraction Tools}
One of the main steps in any information retrieval task is the extraction of the searchable content from the collection of documents. At CDS the most common storage format for publications is PDF. A big part of the harvested documents in CDS comes from the ArXiv pre-prints service and from here, the source code (\LaTeX) of the submissions is available for download. Smaller portions of other documents are stored in other formats like Microsoft Word, OpenOffice Documents.
In practice, the main options from where to do the extraction from, are PDFs and \LaTeX\ files. Also taking into account the previous discussion on the available format for representing mathematics, we focused our attention in tools that output MathML.  \\
A conscious exploration of available software for extracting mathematical content from this two types of documents allowed us to identify different tools with different approaches. 
\LaTeX  files are easier to process and all of the tools implement their own compiler that XML + MathML outputs instead of the typical. Since the original math equation is provided in the document, there is relatively little ambiguity in the equation, and the major issues that the systems struggle with is in the coverage of all the possible commands and packages that are available. \\
For extracting equations from PDF files, the process turns into an Optical Character Recognition (OCR). The reconstruction of the equation has to infer the mathematical structure based on physical positioning and size of the symbols, the length of the lines and so on.
